### 2015-03-19 doc/thesis/system.tex
Fixing typos

Our type system also catches arity mismatch.
To do that, we end the input type of a funcion with type $\Void$
instead of $\Value{*}$.
For instance, let us assume that $f$ has type
$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a top-level return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior not
deterministically, as it makes easier the presentation of this behavior.

>>>>

Our type system also catches arity mismatch.
To do that, we end the input type of a function with type $\Void$
instead of $\Value{*}$.
For instance, let us assume that $f$ has type
$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a top-level return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior in
a not deterministic way, as it makes easier the presentation of this behavior.

### 2015-03-19 doc/thesis/typedlua.tex
Fixing typos

Typed Lua uses local type inference to assign more specific types to
some unannotated declarations.
More precisely, Typed Lua can infer the type of local variables and
the return type of local functions that are not recursive.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.
For local variables, Typed Lua uses the type of the initialization
expression to assign a more specific type to an unannotated local variable.
For local functions, Typed Lua uses the type of the returned expression
to assign a more specific type to an uannotated return type.
This means that we can rewrite the previous example as follows:
\begin{verbatim}
    local function succ (n:integer)
      return n + 1
    end
    local x = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

@@

We can also type variadic functions.
For instance,
\begin{verbatim}
    local function v (...:string):(string*)
      return ...
    end
\end{verbatim}
has type \texttt{(string*) -> (string*)} in both operation modes.
The call \texttt{v()} compiles without any warnings in the default and
strict modes because \texttt{(value*)} and \texttt{()} are consistent
with \texttt{(string*)}.
The call \texttt{v("hello", "world")} compiles without any warnings
in the default and strict modes because \texttt{("hello", "world", nil*)}
and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
The call \texttt{v(...)} compiles withouth any warnings in both
operation modes because the type of the argument list is \texttt{(string*)},
assuming that the vararg expression (\texttt{...}) has type \texttt{string}.

@@

In our example, we are using the refinement of table types to
build the type of the variable \texttt{Shape},
as it should work as our class.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types work.
Finally, the top-level \texttt{return} statement fixes the type of
\texttt{Shape}, because \emph{fixed} table types do not allow width
subtyping, and this behavior allows Typed Lua to use the refinment
of table types to type check single inheritance, as we will see in
this section.

@@

In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has a \emph{fixed} table type then it is safe to
produce an equivalent \emph{open} table type, as \emph{fixed}
table types do not allow hiding fields.
In our example, it is safe to use the type of \texttt{Shape}
to initialize \texttt{Circle}, openning the type of \texttt{Circle}
to allow the refinement of table types to build the type of
the class \texttt{Circle}.

>>>>

Typed Lua uses local type inference to assign more specific types to
some unannotated declarations.
More precisely, Typed Lua can infer the type of local variables and
the return type of local functions that are not recursive.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.
For local variables, Typed Lua uses the type of the initialization
expression to assign a more specific type to an unannotated local variable.
For local functions, Typed Lua uses the type of the returned expression
to assign a more specific type to an unannotated return type.
This means that we can rewrite the previous example as follows:
\begin{verbatim}
    local function succ (n:integer)
      return n + 1
    end
    local x = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

@@

We can also type variadic functions.
For instance,
\begin{verbatim}
    local function v (...:string):(string*)
      return ...
    end
\end{verbatim}
has type \texttt{(string*) -> (string*)} in both operation modes.
The call \texttt{v()} compiles without any warnings in the default and
strict modes because \texttt{(value*)} and \texttt{()} are consistent
with \texttt{(string*)}.
The call \texttt{v("hello", "world")} compiles without any warnings
in the default and strict modes because \texttt{("hello", "world", nil*)}
and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
The call \texttt{v(...)} compiles without any warnings in both
operation modes because the type of the argument list is \texttt{(string*)},
assuming that the vararg expression (\texttt{...}) has type \texttt{string}.

@@

In our example, we are using the refinement of table types to
build the type of the variable \texttt{Shape},
as it should work as our class.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types work.
Finally, the top-level \texttt{return} statement fixes the type of
\texttt{Shape}, because \emph{fixed} table types do not allow width
subtyping, and this behavior allows Typed Lua to use the refinement
of table types to type check single inheritance, as we will see in
this section.

@@

In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has a \emph{fixed} table type then it is safe to
produce an equivalent \emph{open} table type, as \emph{fixed}
table types do not allow hiding fields.
In our example, it is safe to use the type of \texttt{Shape}
to initialize \texttt{Circle}, opening the type of \texttt{Circle}
to allow the refinement of table types to build the type of
the class \texttt{Circle}.
